Transferability of clinical prediction models for early trauma care

### 2019-06-18 ManuscriptMarkdown.Rmd

date: "June 04, 2019"

@@

output: html_document

@@

  **Background:** Trauma results in millions of deaths annually. Trauma care involves many levels of the health care system, and clinical prediction models are often used to triage patients. Previous research has indicated transferring these models might adversely affect model performance.

  **Methods:** Patients age 15 or above registered between the years 2011 and 2016 in the SweTrau registry were included. Three sets of data were created: High and low volume centres, metropolitan and non-metropolitan centres, multi and single centre data. A logistic regression model was developed. Model performance was evaluated in all data sets using mistriage rate, undertriage rate and overtriage rate. Multiple imputation using chained equation was used to handle missing data. Model performance was reported as medians with 95% uncertainty interval across imputed data sets.

@@

Systems to determine the level of trauma can be based on clinical prediction models. Such models are continuously being developed and studied, examining different parameters and different contexts. Models differ in quality and characteristics, but generally perform well predicting survival [@RN45]. Many of these models are developed in a single standardized setting such as a large metropolitan area or a major trauma centre but are then implemented in different settings, such as non-trauma centres.

@@

# Materials and methods
## Study design
We conducted a registry-based cohort study. Sweden has a nationally encompassing register for registering trauma called SweTrau. SweTrau data was used to create clinical prediction models, which were transferred between different contexts to study the effects on mistriage.

@@

## Variables
### Model predictors
The clinical prediction models included the predictors SBP, RR and GCS on arrival to hospital. The rationale for including these three predictors is that they are part of many established clinical prediction models for early trauma care, such as the Revised Trauma Score [@RN35].

@@

### Development and validation sample
Each data set included at least two samples. The samples were then split into two subsamples using a temporal split based on the date of traumatic event. The earlier subsample was the development sample, and the later subsample the validation sample. The development sample contained 70 events (events being patients who died within 30 days of the trauma) and all non-events (non-events being patient survival 30 days past the trauma) during the same time. The rationale for including 70 events is that we needed at least 10 events per free parameter in the logistic regression to obtain stable coefficient estimates [@RN30]. The validation sample instead contained 100 events and at least 100 non-events, which was suggested as the minimum number by Vergouwe in 2005 for external validation samples [@RN37]. See figure 1 for example using the high- and low volume centre data set, with high volume centres being those in the top quartile of number of patients registered.

@@

The minimum sample size of development and validation samples was 140 and 200 respectively. We performed analyses only on data sets for which all samples included at least the minimum number of patients.

### Sequence of analysis
The programming language R was used for all analyses [@RN31], all progress was uploaded to a GitHub repository [@RN59]. Analyses were performed in the sequence of model development, model validation and finally model comparison. These steps were repeated in each data set. Below we used the transfer of a model from a high-volume sample to a low volume sample as an example to describe the complete procedure.

>>>>

date: "June 18, 2019"

@@

output: word_document

@@

  **Background:** Trauma results in millions of deaths annually. Clinical
  prediction models are often used in early trauma care. Previous research has
  indicated transferring these models might adversely affect model performance.

  **Methods:** Patients age 15 or above registered between the years 2011 and
  2016 in the Swedish national trauma registry SweTrau were included. Three sets
  of data were created: High and low volume centres, metropolitan and
  non-metropolitan centres, and multi and single centre data. Clinical
  prediction models based on logistic regression were developed in each set of
  data and transferred within sets. Model performance was evaluated in all data
  sets using mistriage rate, undertriage rate and overtriage rate. Multiple
  imputation using chained equation was used to handle missing data. Model
  performance was reported as medians with 95% uncertainty interval across
  imputed data sets.

@@

Systems to determine the level of trauma can be based on clinical prediction
models.  Models differ in quality and characteristics, but generally perform
well predicting survival [@RN45]. Many of these models are developed in a single
standardized setting such as a large metropolitan area or a major trauma centre
but are then implemented in different settings, such as non-trauma centres.

@@

# Materials and methods
## Study design
We conducted a registry-based cohort study. Sweden has a nationally encompassing
register for registering trauma called SweTrau. SweTrau data was used to create
clinical prediction models, which were transferred between different contexts to
study the effects on mistriage. The study and analysis plans were made
publically available before the research was undertaken (Add reference to github
repo here).

@@

## Variables
### Model predictors
The clinical prediction models included the predictors systolic blood pressure
(SBP), respiratory rate (RR) and Glasgow coma scale (GCS) on arrival to
hospital. The rationale for including these three predictors is that they are
part of many established clinical prediction models for early trauma care, such
as the Revised Trauma Score [@RN35].

@@

### Development and validation sample
Each data set included at least two samples. The samples were then split into
two subsamples using a temporal split based on the date of trauma. The earlier
subsample was the development sample, and the later subsample the validation
sample. The development sample contained 70 events (events being patients who
died within 30 days of the trauma) and all non-events (non-events being patient
survival 30 days past the trauma) during the same time. The rationale for
including 70 events is that we needed at least 10 events per free parameter in
the logistic regression to obtain stable coefficient estimates [@RN30]. The
validation sample instead contained 100 events and at least 100 non-events,
which was suggested as the minimum number by Vergouwe in 2005 for external
validation samples [@RN37]. See figure 1 for example using the high- and low
volume centre data set, with high volume centres being those in the top quartile
of number of patients registered. The minimum sample size of development and
validation samples was 140 and 200 respectively. We performed analyses only on
data sets for which all samples included at least the minimum number of
patients.

@@

### Sequence of analysis
The programming language R was used for all analyses [@RN31], and all code has
been made publically available [@RN59]. Analyses were performed in the sequence
of model development, model validation and finally model comparison. These steps
were repeated in each data set. Below we used the transfer of a model from a
high-volume sample to a low volume sample as an example to describe the complete
procedure.
